---
title: 'Lab 4 Part 2: Intro to time series wrangling and viz'
author: "Katelin Seeto"
date: "2/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(lubridate)# part of tidyverse but doesn't come along automatically
library(tsibble)# for time series tibbles
library(feasts)# for wrangling and working with time series data
library(slider)# for finding moving avg where we specify window
 
```

#### Read in data

```{r}
toolik <- read_csv(here("data", "toolikweather.csv"))  #currently thinks date is character vector, won't plot right
 
```

#### Convert this into tsibble (time series data frame or tibble)

```{r}
toolik_ts <- toolik %>% 
  mutate(date = lubridate::mdy(date)) %>%  #must be in order of existing data, if was year, month, day would use ymd
   as_tsibble(key = NULL, 
             index = date)#index is whatever col has your time series data
  
```

```{r}
ggplot(data= toolik_ts, aes(x = date, y = mean_airtemp))+
  geom_line()

# Now ask some big picture questions: 
#   Is there seasonality, cyclicality, overall trend, notable outliers?
#   In this plot, seasonality is dominant feature


```

#### Use `index_by()` function to aggregate time series info by intervals we specify, is like the `group_by()` function but for time series data 
#### Find averages, max, or other charteristics by intervals we specify

```{r}
# For each year, for each month, what is the mean air temp?
toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>% # is a function of year and month, across entire index
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE))

ggplot(data = toolik_month, aes(x = yr_mo, y = monthly_mean_temp))+ 
  geom_line()+
  facet_wrap(~month(yr_mo, label = TRUE))# Can also split by month across years instead


```

#### Aggregate by week: 
```{r}
toolik_weekly <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% 
  summarize(weekly_airtemp = mean(mean_airtemp, na.rm = TRUE), 
            min_airtemp = min(mean_airtemp))
```

#### Filtering time series 
use `filter_index()` to specify ranges of time series to keep or exclude. 

```{r}
# Filter time series from start of June 2000 to end of October 2001

toolik_ts %>% 
  filter_index("2000-06" ~ "2001-10")

# Filter by day resolution 

toolik_ts %>% 
  filter_index("2000-04-10" ~ "2006-05-15")

# Filter a range and then one random day 

toolik_ts %>% 
  filter_index("2000-04-10" ~ "2006-05-15", "1988-06-01")
```

#### Season plots
Recognizes season and automatically puts series as year, can show how seasonality is changing over time
```{r}
toolik_ts %>% 
  filter(year(date) > 2014) %>% # pulls year from date col and filters for >2014
  gg_season(y = mean_airtemp)


toolik_month %>% 
  gg_season(y = monthly_mean_temp) + 
  theme_light()

```

#### Seasonal subseries plot

```{r}
toolik_month %>% 
gg_subseries(monthly_mean_temp)
```
#### Calculate moving average using the slider package in R

```{r}
#Creating a test vector
set.seed(2021) #psuedo random number generator
test <- rnorm(n = 100,
              mean = 40, 
              sd = 10) # obs pulled from a normal distribution where I want to have 100 total obs, where mean is 40 and sd is 10

slide(test, ~.x, .before = 2, .after = 2) # use slide funct on test, across all values of this vector, make windows so they extend before 2 and after 2, or a window of 5 where the avg would be calculated

```
```{r}

#Calculate rolling average with window size 15 days
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp, 
                                   mean,
                                   .before = 7,
                                   .after = 7))) #the window is usually stored as list so must convert to numeric
 
ggplot()+
  geom_line(data = toolik_ts, aes(x = date, y = mean_airtemp),
            size= 0.2,
            color= "gray") +
  geom_line(data = roll_toolik_15, aes(x = date, y = ma_15d)) + # can graphs as long as data sources are same 
  theme_minimal()
```

#### Create ACF (Autocorrelation Function from feasts)
visual representation of how observations are correlated w/ prior observations on the same variable over specified lags 

```{r}
toolik_ts %>% 
  ACF(mean_airtemp) %>% 
  autoplot()
# Makes sense since days in same month are probably highly correlated

# Try with monthly data
toolik_month %>% 
  ACF(monthly_mean_temp) %>% 
  autoplot()
# Now see clear and reflected seasonality, in our ACF
# This means we can see on x axis there is a lag in months, observations that are 12mo apart are strongly positively correlated, and obs that are 6mo apart (opposite seasons) are negatively correlated, and not much dampening over years 

```

#### Decomposition by STL (seasonal trend and lowest composition)

```{r}
toolik_dec <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf)))# model with the STL model, start with monthly mean temp modeled as a function of the different seasons for a period of time specified by Inf, but is all stored as a mondel.... not useful

components(toolik_dec) %>% #pulls the components out of model to plot
  autoplot()

```



